Using reference models in feature selection problems \vspace{.1in}

### 2019-07-17 paper/ref_approach.tex
conclusion and typo edits

In statistical applications, one of the main steps in the modeling
workflow is covariate or feature selection, which is a special case of
model reduction. Variable selection may have multiple goals. 
First, if the covariates themselves are of interest,
we can use variable selection as an inferential tool to find the entire 
collection of variables which have some predictive information about the target variable.
Second, as simpler models come with the advantages of reduced
measurement costs and improved interpretability, we may be interested in
finding the minimal subset of covariates which
provides good predictive performance (or good balance between simplicity and
predictive performance). 
Whenever the predictive capability is guiding the selection, the true data generation
mechanism of future data can be approximated either by using the observed data directly or
alternatively by using predictions from a \emph{reference model}.

In data based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward 
regression \citep{venables2013modern,harrell2015regression}, the observed empirical data distribution is utilized as a
proxy of future data usually in combination with cross-validation
or information criteria to provide estimates for out-of-sample
predictive performance.
In contrast, reference model based methods approximate the future data generation
mechanism using the predictive distribution of a reference model,
which can be, for example, a full-encompassing model including all
covariates. The reference
model approach has been used for long time in Bayesian statistics
since the seminal work of \citet{paper:reference_lindley}. For more
historical references, see the review by \citet{vehtari2012survey}, and for
most recent methodological development see 
\citet{paper:projpred}.
Reference models have been also used in non-Bayesian contexts,
where \cite{harrell2015regression} describes them as full models that
can be thought of as a ``gold standard'' (for a given application).
For example,
\cite{faraggi2001understanding} deal with the necessity of identifying
interpretable risk groups in the context of survival data using neural
networks, which typically perform very well in terms of prediction,
but whose covariates are difficult to be understood in terms of
relevance. 
\cite{paul2008preconditioning}, using the term preconditioning,
explore approximating models fitting Lasso or stepwise regression
against consistent estimates $\hat{y}$ of a reference model instead of
the observed responses $y$.

@@

As a motivational example to the implications of using a reference model in feature selection problems, we show its application in the form of the projection predictive approach, which uses the reference model to project the posterior samples to the submodels \citep{paper:original_proj}. The \texttt{R}-package \texttt{projpred} (available at \url{https://CRAN.R-project.org/package=projpred}) implements the projection in case of submodels in the family of the generalised linear models and, additionally, it provides a framework to choose the optimal submodel size through a predictive utility comparison between the submodels and the reference model. 
\\
We now summarise the workflow of the projection predictive approach in the particular case of the draw-by-draw projection \cite[original formulation by][]{paper:original_proj}, following \cite{paper:projpred}. Suppose we have observed $n$ statistical units with target values $\{y_{i}\}_{i=1}^{n}$ and a set of observed covariates for which we want to obtain a minimally relevant subset. Than, the main steps are the following:
\begin{enumerate}
\item Devise and fit a reference model. Let $\{\boldsymbol{\theta}_{*}^{s}\}_{s=1}^{S}$ be the set of $S$ posterior samples of the reference model's parameters.
\item Rank the covariates according to their relevance using some heuristics and consider as candidate submodels only those which preserve this order,  starting from including only the highest ranked covariate (the submodels are then naturally identified by their model size). This is not strictly necessary but reduces the number of submodels to consider in the following steps.
\item For each submodel, project the reference model's samples as follows:
\begin{equation}\label{eq:draw_by_draw}
\boldsymbol{\theta}_{\perp}^{s} = \underset{\boldsymbol{\theta^{s}}\in\Theta}{\text{argmin}} \frac{1}{n}\sum_{i=1}^{n}\text{KL} \left( p(\tilde{y}_{i}|\boldsymbol{\theta}_{*}^{s})\;||\;q(\tilde{y}_{i}|\boldsymbol{\theta}^{s}) \right) \quad s=1,..,S
\end{equation}
where $p(\tilde{y}_{i}|\boldsymbol{\theta}_{*}^{s})$ stands for the predictive distribution of the reference model with parameters fixed at the posterior sample $\boldsymbol{\theta}_{*}^{s}$ and conditioning on all the covariate values related to the statistical unit (identified by the subscript $i$), whereas $q(\tilde{y}_{i}|\boldsymbol{\theta}^{s})$ is the predictive distribution of the submodel. The projected samples $\boldsymbol{\theta}_{\perp}^{s}$ then correspond to the submodel's parameter distribution.
\item For each submodel (size), test the predictive performance for instance via cross-valdiation.
\item Choose the smallest submodel (size) that is sufficiently close to the reference model's predictive utility score.
\end{enumerate}

@@

\cite{paper:bodyfat} report results using steplm with a significance level of 0.157 with AIC selection \citep{akaike1974new}, fixing abdomen and height to be always included in the model. Here, we implement steplm via AIC without fixing any variable. For selection via projpred, the reference model includes all the covariates, we regularize model complexity with a regularised horseshoe prior on the covariate coefficients \citep{paper:rhs}, submodels are explored using forward search, and the predictive utility is the expected log-predictive density (elpd) estimated using PSIS-LOO \citep{paper:psis_loo}. The experiments are implemented in \texttt{R} \citep{Rcore2018}. The steplm approach is carried out combining the \texttt{step} and \texttt{lm} functions, whereas the selection via projpred is done with the \texttt{projpred} package. In order to automatise the latter, we select the submodel as the one with the smallest size which has an elpd score higher (i.e., better) than the reference model score with probability of at least 5\%. The inclusion frequencies of each feature in the final model after 100 bootstrap samples are depicted in Figure \ref{fig:inclusion_frequencies}. The projpred approach has only two variables, 'abdomen' and 'weight', with inclusion frequencies above 50\% ('abdomen' is the only included always), the third most included is `wrist' at 44\%, and the fourth is `height' at 35\%, while steplm has seven features above 50\%. Such a lower stability of steplm can be observed also in the bootstrap model selection frequencies reported in Table \ref{tab:model_frequencies}. For exaple, the first five selected models have a cumulative frequency of 76\% with projpred, but only of 14\% with steplm. In addition, we note that the size of the selected model is much smaller when using projpred.

@@

Table \ref{tab:model_performances} reports the predictive performances, in terms of root mean square error, of the full model and the corresponding selected submodel using both projpred and steplm. We also repeat the selection after adding truly uncorrelated (i.e., irrelevant) features up to a total of 100 covariates. For this experiment, we report both the size of the selected submodel and the number of selected irrelevant features. Whilst, with the original data, we do not observe any gain using projpred in terms of predictive performance, what we do observe is a smaller size for the selected submodel, at the cost of a higher computation time compared to steplm. However, when additional irrelevant features are added, projpred is able to notably improve the predictive performance keeping the same submodel size, whereas steplm selects a much larger submodel also including a large number of irrelvant features (13 irrelevant out of a total of 19 selected features).

@@

In this section, we demonstrate the benefits of reference model approaches using the normal means problem as a benchmark. The normal means problem consists of estimating the (usually sparse) vector of means of a vector of normally distributed observations. Let us denote the dimensionality of the vector of means by $p$ and let $\{z_{j}\}_{j=1}^{p}$ be the vector of observations of the random variables $\{Z_{j}\}_{j=1}^{p}$. In the normal means problem, the task is to estimate the latent variables $\{\theta_{j}\}_{j=1}^{p}$ of the following model:
\
\begin{equation}\label{eq:normal_means_problem}
Z_{j}|\theta_{j},\sigma\overset{ind}{\sim}N(\theta_{j},\sigma^{2}) \quad j=1,..,p
\end{equation}
Note that it is equivalent to a linear regression where the design matrix is the identity matrix with the number of features being equal to the number of observations. This formulation can be found in practice at various places, 
for example, in the analysis of microarray data, where a large set of genes are tested in two groups of patients labeled as positive or negative to some disease \citep{paper:efron,efron2012large}. The objective of the analysis is to select the subset of genes statistically relevant to the disease. One common way to proceed is to compute the two-samples $t$-statistic for every gene separately.
After normalizing these statistics, they then become the data, $Z$, in the normal means problem \eqref{eq:normal_means_problem}. For further details see the examples in \cite{paper:efron, efron2012large}.

In our examples, we retrieve the normal means problem from the sample correlations between the target and the covariates using the Fisher $z$-transformation \citep{hawkins1989using}. Suppose we have a continuous target random variable $Y$, a set of $p$ continuous covariates $\{X_{j}\}_{j=1}^{p}$, and let us denote $\rho_{j}=Cor(Y,X_{j})$. Further, suppose we have observed $n$ statistical units and define $r_{j}$ the sample correlation between the observations of the target variables $\{y_{i}\}_{i=1}^{n}$ and the $j$-th covariate $\{x_{ij}\}_{i=1}^{n}$. Finally, we refer to the Fisher $z$-transformation function $\text{tanh}^{-1}(\cdot)$ as $T_{F}(\cdot)$. Assuming each pair $(Y,X_{j})$ to be bivariate normally distributed, the corresponding transformed correlations are approximately normally distributed as well with known variance:
\
\begin{equation} \label{eq:fisher_transformation}
T_{F}(r_{j})\overset{ind}{\sim} N(T_{F}(\rho_{j}),\frac{1}{n-3}) \quad j=1,..,p
\end{equation}
Therefore, rescaling the quantities $T_{F}(r_{j})$ by $\sqrt{n-3}$ and referring to the results as variables $z_{j}$, we again find the formulation \eqref{eq:normal_means_problem} of the normal means problem, this time with unit variance:
\
\begin{equation} \label{eq:normal_means_problem2}
Z_{j}|\theta_{j}\overset{ind}{\sim}N(\theta_{j},1) \quad j=1,..,p
\end{equation}
Note that in thise cse the quantities of interest $\theta_{j}$ are equal to $\sqrt{n-3}\,T_{F}(\rho_{j})$. There are different ways to proceed with the selection in the normal means problem. Here, we consider several state-of-the-art approaches primarily from Bayesian statistics.

@@

Figure \ref{fig:bodyfat_sensitivity_vs_fdr} shows the sensitivity against the false discovery rate. Since we do not have a ground truth available with regard the original covariates of the data, we believe it is reasonable to consider all of them relevant, at least at some degree. The artificially added featues as of course irrelvant by construction. In almost all the bootstrapped subsamples, the reference model improves the seleciton both in terms of sensitivity and false discovery rate. When $n=50$, we observe a worsening in terms of false discovery rate, yet by a lower amount compared to the gain in sensitivity. Again, we observe that the benefits are more evident as the selection becomes more challenging (i.e., lower number of observations). Figure \ref{fig:bodyfat_stability} shows the stability results, still using the measure provided by \cite{paper:stability}. The benefits of the reference model are here marginal: small improvements can be observed, but overall not very noteworthy.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \caption{Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelvant features on the left and the out-of-sample RMSE on the right-hand side after 100 bootstrap samples.\\}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}

@@

In this paper, we discussed the importance of using a reference model to improve feature selection, or more generally, model reduction. We want to highlight that the goal of this paper is not to provide a specific procedure to carry out the selection, but rather to explain and motivate the general use and benefits of a reference model. 
We have seen how the reference model acts as an approximation of the data generation mechanism through its predictive distribution. Such approximation is generally less noisy than the sample estimation available purely from the observed data, leading to the main benefits of the reference model approach. In our comparisons, we have analysed the effect of a reference model in the form of a filter on the observed target values on top of different widely used variable selection methods. Overall, using a reference model lead to more accurate and stable selection results independently of the specific selection method. These benefits apply to a large family of different approaches all involving a reference model in one way or the other. Some of these approches have been present in the literature for several years (cite) but often without a clear explanation of \emph{why} they are actually favorable and how they connect to other (as it turns out) related approaches. We hope that the present paper can fill some of these gaps by providing a unifying framework and understanding of reference models.

>>>>

In statistical applications, one of the main steps in the modelling
workflow is covariate or feature selection, which is a special case of
model reduction. Variable selection may have multiple goals. 
First, if the covariates themselves are of interest,
we can use variable selection as an inferential tool to find the entire 
collection of variables which have some predictive information about the target variable.
Second, as simpler models come with the advantages of reduced
measurement costs and improved interpretability, we may be interested in
finding the minimal subset of covariates which
provides good predictive performance (or good balance between simplicity and
predictive performance). 
Whenever the predictive capability is guiding the selection, the true data generation
mechanism of future data can be approximated either by using the observed data directly or
alternatively by using predictions from a \emph{reference model}.

In data based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward 
regression \citep{venables2013modern,harrell2015regression}, the observed empirical data distribution is utilised as a
proxy of future data usually in combination with cross-validation
or information criteria to provide estimates for out-of-sample
predictive performance.
In contrast, reference model based methods approximate the future data generation
mechanism using the predictive distribution of a reference model,
which can be, for example, a full-encompassing model including all
covariates. The reference
model approach has been used for long time in Bayesian statistics
since the seminal work of \citet{paper:reference_lindley}. For more
historical references, see the review by \citet{vehtari2012survey}, and for
most recent methodological development see 
\citet{paper:projpred}.
Reference models have been also used in non-Bayesian contexts,
where \cite{harrell2015regression} describes them as full models that
can be thought of as a ``gold standard'' (for a given application).
For example,
\cite{faraggi2001understanding} deal with the necessity of identifying
interpretable risk groups in the context of survival data using neural
networks, which typically perform very well in terms of prediction,
but whose covariates are difficult to be understood in terms of
relevance. 
\cite{paul2008preconditioning}, using the term preconditioning,
explore approximating models fitting Lasso or stepwise regression
against consistent estimates $\hat{y}$ of a reference model instead of
the observed responses $y$.

@@

As a motivational example to the implications of using a reference model in feature selection problems, we show its application in the form of the projection predictive approach, which uses the reference model to project the posterior samples to the submodels \citep{paper:original_proj}. The \texttt{R}-package \texttt{projpred} (available at \url{https://CRAN.R-project.org/package=projpred}) implements the projection in case of submodels in the family of the generalised linear models and, additionally, it provides a framework to choose the optimal submodel size through a predictive utility comparison between the submodels and the reference model. 
\\
We now summarise the workflow of the projection predictive approach in the particular case of the draw-by-draw projection \cite[original formulation by][]{paper:original_proj}, following \cite{paper:projpred}. Suppose we have observed $n$ statistical units with target values $\{y_{i}\}_{i=1}^{n}$ and a set of observed covariates for which we want to obtain a minimally relevant subset. Than, the main steps are the following:
\begin{enumerate}
\item Devise and fit a reference model. Let $\{\boldsymbol{\theta}_{*}^{s}\}_{s=1}^{S}$ be the set of $S$ posterior samples of the reference model's parameters.
\item Rank the covariates according to their relevance using some heuristics and consider as candidate submodels only those which preserve this order,  starting from including only the highest ranked covariate (the submodels are then naturally identified by their model size). This is not strictly necessary but reduces the number of submodels to consider in the following steps.
\item For each submodel, project the reference model's samples as follows:
\begin{equation}\label{eq:draw_by_draw}
\boldsymbol{\theta}_{\perp}^{s} = \underset{\boldsymbol{\theta^{s}}\in\Theta}{\text{argmin}} \frac{1}{n}\sum_{i=1}^{n}\text{KL} \left( p(\tilde{y}_{i}|\boldsymbol{\theta}_{*}^{s})\;||\;q(\tilde{y}_{i}|\boldsymbol{\theta}^{s}) \right) \quad s=1,..,S
\end{equation}
where $p(\tilde{y}_{i}|\boldsymbol{\theta}_{*}^{s})$ stands for the predictive distribution of the reference model with parameters fixed at the posterior sample $\boldsymbol{\theta}_{*}^{s}$ and conditioning on all the covariate values related to the statistical unit (identified by the subscript $i$), whereas $q(\tilde{y}_{i}|\boldsymbol{\theta}^{s})$ is the predictive distribution of the submodel. The projected samples $\boldsymbol{\theta}_{\perp}^{s}$ then correspond to the submodel's parameter distribution.
\item For each submodel (size), test the predictive performance for instance via cross-validation.
\item Choose the smallest submodel (size) that is sufficiently close to the reference model's predictive utility score.
\end{enumerate}

@@

\cite{paper:bodyfat} report results using steplm with a significance level of 0.157 with AIC selection \citep{akaike1974new}, fixing abdomen and height to be always included in the model. Here, we implement steplm via AIC without fixing any variable. For selection via projpred, the reference model includes all the covariates, we regularise model complexity with a regularised horseshoe prior on the covariate coefficients \citep{paper:rhs}, submodels are explored using forward search, and the predictive utility is the expected log-predictive density (elpd) estimated using PSIS-LOO \citep{paper:psis_loo}. The experiments are implemented in \texttt{R} \citep{Rcore2018}. The steplm approach is carried out combining the \texttt{step} and \texttt{lm} functions, whereas the selection via projpred is done with the \texttt{projpred} package. In order to automatise the latter, we select the submodel as the one with the smallest size which has an elpd score higher (i.e., better) than the reference model score with probability of at least 5\%. The inclusion frequencies of each feature in the final model after 100 bootstrap samples are depicted in Figure \ref{fig:inclusion_frequencies}. The projpred approach has only two variables, 'abdomen' and 'weight', with inclusion frequencies above 50\% ('abdomen' is the only included always), the third most included is `wrist' at 44\%, and the fourth is `height' at 35\%, while steplm has seven features above 50\%. Such a lower stability of steplm can be observed also in the bootstrap model selection frequencies reported in Table \ref{tab:model_frequencies}. For example, the first five selected models have a cumulative frequency of 76\% with projpred, but only of 14\% with steplm. In addition, we note that the size of the selected model is much smaller when using projpred.

@@

Table \ref{tab:model_performances} reports the predictive performances, in terms of root mean square error, of the full model and the corresponding selected submodel using both projpred and steplm. We also repeat the selection after adding truly uncorrelated (i.e., irrelevant) features up to a total of 100 covariates. For this experiment, we report both the size of the selected submodel and the number of selected irrelevant features. Whilst, with the original data, we do not observe any gain using projpred in terms of predictive performance, what we do observe is a smaller size for the selected submodel, at the cost of a higher computation time compared to steplm. However, when additional irrelevant features are added, projpred is able to notably improve the predictive performance keeping the same submodel size, whereas steplm selects a much larger submodel also including a large number of irrelevant features (13 irrelevant out of a total of 19 selected features).

@@

In this section, we demonstrate the benefits of reference model approaches using the normal means problem as a benchmark. The normal means problem consists of estimating the (usually sparse) vector of means of a vector of normally distributed observations. Let us denote the dimensionality of the vector of means by $p$ and let $\{z_{j}\}_{j=1}^{p}$ be the vector of observations of the random variables $\{Z_{j}\}_{j=1}^{p}$. In the normal means problem, the task is to estimate the latent variables $\{\theta_{j}\}_{j=1}^{p}$ of the following model:
\
\begin{equation}\label{eq:normal_means_problem}
Z_{j}|\theta_{j},\sigma\overset{ind}{\sim}N(\theta_{j},\sigma^{2}) \quad j=1,..,p
\end{equation}
Note that it is equivalent to a linear regression where the design matrix is the identity matrix with the number of features being equal to the number of observations. This formulation can be found in practice at various places, 
for example, in the analysis of microarray data, where a large set of genes are tested in two groups of patients labeled as positive or negative to some disease \citep{paper:efron,efron2012large}. The objective of the analysis is to select the subset of genes statistically relevant to the disease. One common way to proceed is to compute the two-samples $t$-statistic for every gene separately.
After normalising these statistics, they then become the data, $Z$, in the normal means problem \eqref{eq:normal_means_problem}. For further details see the examples in \cite{paper:efron, efron2012large}.

In our examples, we retrieve the normal means problem from the sample correlations between the target and the covariates using the Fisher $z$-transformation \citep{hawkins1989using}. Suppose we have a continuous target random variable $Y$, a set of $p$ continuous covariates $\{X_{j}\}_{j=1}^{p}$, and let us denote $\rho_{j}=Cor(Y,X_{j})$. Further, suppose we have observed $n$ statistical units and define $r_{j}$ the sample correlation between the observations of the target variables $\{y_{i}\}_{i=1}^{n}$ and the $j$-th covariate $\{x_{ij}\}_{i=1}^{n}$. Finally, we refer to the Fisher $z$-transformation function $\text{tanh}^{-1}(\cdot)$ as $T_{F}(\cdot)$. Assuming each pair $(Y,X_{j})$ to be bivariate normally distributed, the corresponding transformed correlations are approximately normally distributed as well with known variance:
\
\begin{equation} \label{eq:fisher_transformation}
T_{F}(r_{j})\overset{ind}{\sim} N(T_{F}(\rho_{j}),\frac{1}{n-3}) \quad j=1,..,p
\end{equation}
Therefore, rescaling the quantities $T_{F}(r_{j})$ by $\sqrt{n-3}$ and referring to the results as variables $z_{j}$, we again find the formulation \eqref{eq:normal_means_problem} of the normal means problem, this time with unit variance:
\
\begin{equation} \label{eq:normal_means_problem2}
Z_{j}|\theta_{j}\overset{ind}{\sim}N(\theta_{j},1) \quad j=1,..,p
\end{equation}
Note that in this case the quantities of interest $\theta_{j}$ are equal to $\sqrt{n-3}\,T_{F}(\rho_{j})$. There are different ways to proceed with the selection in the normal means problem. Here, we consider several state-of-the-art approaches primarily from Bayesian statistics.

@@

Figure \ref{fig:bodyfat_sensitivity_vs_fdr} shows the sensitivity against the false discovery rate. Since we do not have a ground truth available with regard the original covariates of the data, we believe it is reasonable to consider all of them relevant, at least at some degree. The artificially added features as of course irrelevant by construction. In almost all the bootstrapped subsamples, the reference model improves the selection both in terms of sensitivity and false discovery rate. When $n=50$, we observe a worsening in terms of false discovery rate, yet by a lower amount compared to the gain in sensitivity. Again, we observe that the benefits are more evident as the selection becomes more challenging (i.e., lower number of observations). Figure \ref{fig:bodyfat_stability} shows the stability results, still using the measure provided by \cite{paper:stability}. The benefits of the reference model are here marginal: small improvements can be observed, but overall not very noteworthy.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \caption{Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant features on the left and the out-of-sample RMSE on the right-hand side after 100 bootstrap samples.\\}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}

@@

In this paper, we discussed the importance of using a reference model to improve feature selection, or more generally, model reduction. We want to highlight that the goal of this paper is not to provide a specific procedure to carry out the selection, but rather to explain and motivate the general use and benefits of a reference model. 
We have seen how the reference model acts as an approximation of the data generation mechanism through its predictive distribution. Such approximation is generally less noisy than the sample estimation available purely from the observed data, leading to the main benefits of the reference model approach. In our comparisons, we have analysed the effect of a reference model in the form of a filter on the observed target values on top of different widely used variable selection methods. Overall, using a reference model lead to more accurate and stable selection results independently of the specific selection method. These benefits apply to a large family of different approaches all involving a reference model in one way or the other. Some of these approaches have been present in the literature for several years \cite[e.g. see][]{vehtari2012survey,paper:projpred} but often without a clear explanation of \emph{why} they are actually favourable and how they connect to other (as it turns out) related approaches. We hope that the present paper can fill some of these gaps by providing a unifying framework and understanding of reference models.
